Parsimonious Representation of Nonlinear Dynamical Systems Through Manifold Learning: A Chemotaxis Case Study

### 2015-05-20 chemotaxis/paper/draft1.tex
updated chemotaxis with Yannis' edits

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{cite}
%\usepackage{graphics}
%\usepackage{epsf}
%\usepackage{psfig}
%\usepackage{float}
\usepackage{graphicx}
\usepackage{epstopdf}
%\usepackage[small]{caption2}
%\usepackage{bm}%
%\usepackage{multirow}
%\usepackage{subfigure}
%\usepackage{wrapfig}
%\usepackage{color}
%\usepackage{dsfont}
%\usepackage{amsthm}
%\usepackage{algorithm}
%\usepackage{subfigure}
\usepackage{subcaption}

@@

\address[princeton]{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08540, USA}
\address[technion]{Technion - Israel Institute of Technology, Haifa, 3200003, Israel}
\address[yale]{Department of Mathematics, Yale University, New Haven, CT, 06520, USA}
\address[princetonpacm]{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, 08540, USA}
%
\cortext[cor1]{Corresponding author}

\begin{abstract}
%
Nonlinear manifold learning algorithms, such as diffusion maps, have been fruitfully
applied in recent years to the analysis of large and complex data sets.
%
However, such algorithms still encounter challenges when faced with real data.
%
One such 
challenge is the existence of ``repeated eigendirections," which obscures the detection of the 
true dimensionality of the underlying manifold and arises when several embedding coordinates parametrize
the same direction in the intrinsic geometry of the data set. 
%
We propose an algorithm, based on local linear regression, 
to automatically detect coordinates corresponding to repeated eigendirections.
%
We construct a more parsimonious embedding using only the eigenvectors corresponding to unique eigendirections, and we show that this reduced diffusion maps embedding induces 
a metric which is equivalent to the standard diffusion distance.
%
We first demonstrate the utility and flexibility of our approach on synthetic data sets.
%
We then apply our algorithm to data collected from a stochastic simulation of cellular chemotaxis, where our approach for factoring out repeated eigendirections allows us to detect changes in dynamical behavior and the underlying
intrinsic system dimensionality directly from data.
%
\end{abstract}

@@

%Due to the complexity and range of microscale behaviors possible in dynamical systems,
%we use manifold learning, a class of nonlinear data-driven techniques, to analyze data.
%
Several manifold learning algorithms obtain parametrizations of the data through
the spectral analysis of a Laplace operator \cite{Belkin2003,coifman2005geometric,coifman2006geometric,singer2008non}.
%
The data is then embedded in a new low-dimensional coordinate system
given by the eigenvectors of this operator.
%
The premise is that these coordinates, obtained in a data-driven manner, are the right macroscopic ``observables",
the variables which parametrize the macroscale dynamical behavior of the system, thus enabling the construction of a reduced macroscale model.

Such manifold learning algorithms were initially applied to synthetic data sets, to illustrate
their geometric properties and flexibility \cite{coifman2005geometric, nadler2006diffusion}.
%
More recently they have been applied to experimental and simulation data,
enabled by advances in data representation (observers) and
metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally, rohrdanz2011determination}.
%
A nontrivial shortcoming of these methods is that, from a geometric perspective,
not all eigenvectors are guaranteed to parametrize unique directions within the data;
some eigenvectors are ``repeated eigendirections'' which describe the same coordinate in the intrinsic geometry
of the data.
%
Identifying these eigenvectors is critical for obtaining a
parametrization of the system which captures the true dimensionality of the macroscale dynamics.
%
This task is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{gerber2007robust}.

In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression \cite{wasserman2006all}.
%
We first demonstrate and validate our algorithm on synthetic examples, where a closed-form solution for the
eigenvectors and eigenvalues of the Laplace operator is known.
%
We then consider a simulation data set from a stochastic dynamical system modeling
cellular chemotaxis \cite{othmer1988models}.
%
Recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections,
provide a data analysis pipeline which successfully analyzes this simulation in a purely data-driven manner.
%
We will show that this pipeline allows us to detect changes in the dimensionality of the underlying macroscale dynamics, which are
related to changes in the regime/mode of the system.

@@

Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ state observations sampled from an evolving autonomous dynamical system.
%
We assume the $n$-dimensional observations $z(i)$ lie on a $d$-dimensional manifold $\mathcal{M}_d$, where $d < n$.
%
We therefore consider the problem of parametrizing the continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$ from data.
%
For linear hyperplanes, the principal axes parametrize the manifold.
%
However, in the case when the manifold is nonlinear, a set of appropriate coordinates is not readily apparent.
%
By using a particular manifold learning technique based on the construction of a Laplace operator, namely, diffusion maps, we will show how to extract a $d$-dimensional parametrization of the observations which is consistent with the geometry of the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_cnts1}
\includegraphics[width=\textwidth]{strip_cnts2}
\caption{}
\label{subfig:strip_efuncs}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\includegraphics[width=\textwidth]{strip_discrete4}
\caption{}
\label{subfig:strip_evecs_uniform}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_nonuniform1}
\includegraphics[width=\textwidth]{strip_nonuniform2}
\caption{}
\label{subfig:strip_evecs_nonuniform}
\end{subfigure}
%
\caption{(a) Two-dimensional continuous strip colored by the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$. (b) Two-dimensional strip with uniform sampling colored by the first and fourth (non-trivial) eigenvectors of the discrete Laplacian. (c) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the first and third (non-trivial) eigenvectors of the discrete Laplacian. Note that in all cases we uncover parametrizations which are one-to-one with $z_1$ and $z_2$.}
\end{figure}

In recent years it has been noted that the eigenfunctions of the \emph{continuous}
Laplace-Beltrami operator provide ``good" coordinates for a manifold \cite{jones2008}.
%
We start by considering such a continuous setting, where rigorous analysis is possible for specific examples.
%
Using such an example, we will demonstrate that the existence of repeated eigendirections is
inherent to the manifold learning setup based on Laplace operators,
and that the identification of the unique eigendirections is nontrivial.
%
The fact that these challenges arise in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for data analysis.

To illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$.
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}.
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and
$\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates,
respectively, and therefore provide a parametrization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide an estimate of the
relative magnitude of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases,
the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases
(this will be discussed further in Section~\ref{sec:relative_lengths}).
%
The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper.
%
First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions,
and a proper parameterization of the manifold is not necessarily given by the $d$ eigenfunctions associated with the
smallest $d$ eigenvalues.
% TODO: address this in later section
%
Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any
additional directions intrinsic to the strip geometry; we will refer to these as ``repeated eigendirections,''
and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
%
Clearly, this problem of repeated eigendirections also arises for more complex manifolds of more than two variables.
%
We note that, although the eigenfunctions can only be written analytically for very special cases,
it has been observed empirically that they often provide appropriate coordinates
to usefully parametrize more complex, nonlinear manifolds.
%

@@

Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we first construct the weight matrix $\mathbf{W} \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
\mathbf{W}_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations.
%
The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{rohrdanz2011determination, coifman2008graph};
we often take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $\mathbf{D} \in \mathbb{R}^{m \times m}$, with $\mathbf{D}_{ii} = \sum_j \mathbf{W}_{ij}$,
and form the matrix $\widetilde{\mathbf{W}} = \mathbf{D}^{-\alpha} \mathbf{W} \mathbf{D}^{-\alpha}$, where $0 < \alpha < 1$.
%
Next, we construct the diagonal matrix $\widetilde{\mathbf{D}} \in \mathbb{R}^{m \times m}$, with $\widetilde{\mathbf{D}}_{ii} = \sum_j \widetilde{\mathbf{W}}_{ij}$, and the matrix $\mathbf{A}  = \widetilde{\mathbf{D}}^{-1} \widetilde{\mathbf{W}}.$

If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then, for $\epsilon \rightarrow 0$ and $m \rightarrow \infty$ (with the appropriate rates), the discrete matrix converges to the following continuous limit operators with Neumann boundary conditions (as discussed in the previous section) \cite{coifman2006geometric}
\begin{align}
\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - 2\nabla U \cdot \nabla \phi, &&\alpha = 0 \\
\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, &&\alpha = 1/2 \\
\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi, &&\alpha = 1
\end{align}
where $U = - \log q$.
%
The different limit operators, depending on the choice of $\alpha$, imply that nonuniform
sampling on the manifold may have different effects.
%
In particular, by setting $\alpha=1$, one can factor out the density effects in the
weight matrix, and the discrete Laplacian matrix approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$.
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $\mathbf{A}$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $\mathbf{A}$ are related to the eigenvalues of the continuous operator by
\begin{equation} \label{eq:evals_relationship}
\mu_k = \exp \left( -\frac{\epsilon^2}{4} \tilde{\mu}_{k_1, k_2}  \right).
\end{equation}

@@

From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$
provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.
%
From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of
the corresponding eigenvalues implies that these two eigenvectors are guaranteed to
appear before any coupled or repeated eigendirections.
%
However, these eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors,
as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
%
We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold are unknown and cannot easily be obtained from the coordinates of the original data, so that identifying which eigenvectors correspond to unique eigendirections is nontrivial.

@@

Common practice is to order the eigenvectors by the magnitude of the corresponding eigenvalues,
assuming that the leading eigenvectors provide a parametrization of the underlying manifold.
%
However, as discussed in the previous section, some eigenvectors are higher harmonics of
previous eigenvectors and do not describe new directions in the data set \cite{gerber2007robust}.
%
For the case where $\mathcal{M}_d$ is a $2$-dimensional strip with edge lengths $L_1  > L_2$, recall
that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left(  {\pi z_1}/{L_1} \right)$ and
$\tilde{\phi}_{0,1} = \cos \left(  {\pi z_2}/{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to
correspond to the two smallest (non-trivial) eigenvalues (the smallest eigenvalue will always be $\tilde{\mu}_{0,0} = 0$ and
correspond to a constant eigenfunction; these eigenvalues are related to the eigenvalues $\mu_k$ of the discrete
operator via \eqref{eq:evals_relationship}).
%
In fact, if $L_1 > 2 L_2$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the
second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues)
will be a ``repeated eigendirection" of the first and still parametrize $z_1$ (see Figure~\ref{fig:strip_harmonics}).
%
Automatic detection of which eigenvectors are harmonics of each other would be clearly useful.
%
Utilizing only the eigenvectors $\phi_i$ which correspond to unique eigendirections yields
the most parsimonious representation of the data.
%
We will show how we can automatically detect these eigenvectors to obtain a meaningful representation of the data,
and that using only these eigenvectors yields a reduced embedding which is equivalent to the standard diffusion maps embedding.
%
Furthermore, the corresponding eigenvalues provide us with an estimate of the relative lengths of the data set along these unique eigendirections.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\end{subfigure}
%
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete2}
\end{subfigure}
%
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete3}
\end{subfigure}
%
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete4}
\end{subfigure}
\caption{ Two-dimensional strip with uniform sampling colored by the first four (non-trivial) eigenvectors from diffusion maps. Note that the first and fourth eigenvectors are one-to-one with $z_1$ and $z_2$, respectively. However, the second and third eigenvectors are higher harmonics of the first eigenvector and do not capture any additional structure within the data set. }
\label{fig:strip_harmonics}
\end{figure}

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{cite}
%\usepackage{graphics}
%\usepackage{epsf}
%\usepackage{psfig}
%\usepackage{float}
\usepackage{graphicx}
\usepackage{epstopdf}
%\usepackage[small]{caption2}
%\usepackage{bm}%
%\usepackage{multirow}
%\usepackage{subfigure}
%\usepackage{wrapfig}
%\usepackage{color}
%\usepackage{dsfont}
%\usepackage{amsthm}
%\usepackage{algorithm}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{hyperref}

@@

\address[princeton]{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08540, USA}
\address[technion]{Department of Electrical Engineering, Technion - Israel Institute of Technology, Haifa, 3200003, Israel}
\address[yale]{Department of Mathematics, Yale University, New Haven, CT, 06520, USA}
\address[princetonpacm]{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, 08540, USA}
%
\cortext[cor1]{Corresponding author}

\begin{abstract}
%
Nonlinear manifold learning algorithms, such as diffusion maps, have been fruitfully
applied in recent years to the analysis of large and complex data sets.
%
However, such algorithms still encounter challenges when faced with real data.
%
One such 
challenge is the existence of ``repeated eigendirections," which obscures the detection of the 
true dimensionality of the underlying manifold and arises when several embedding coordinates parametrize
the same direction in the intrinsic geometry of the data set. 
%
We propose an algorithm, based on local linear regression, 
to automatically detect coordinates corresponding to repeated eigendirections.
%
We construct a more parsimonious embedding using only the eigenvectors corresponding to unique eigendirections, and we show that this reduced diffusion maps embedding induces 
a metric which is equivalent to the standard diffusion distance.
%
We first demonstrate the utility and flexibility of our approach on synthetic data sets.
%
We then apply our algorithm to data collected from a stochastic model of cellular chemotaxis, where our approach for factoring out repeated eigendirections allows us to detect changes in dynamical behavior and the underlying
intrinsic system dimensionality directly from data.
%
\end{abstract}

@@

%Due to the complexity and range of microscale behaviors possible in dynamical systems,
%we use manifold learning, a class of nonlinear data-driven techniques, to analyze data.
%
Several manifold learning algorithms obtain parametrizations of the data through
the spectral analysis of a Laplace operator \cite{Belkin2003,coifman2005geometric,coifman2006geometric,singer2008non}.
%
The data is then embedded in a new low-dimensional coordinate system
given by the eigenvectors of this operator.
%
The premise is that these coordinates, obtained in a data-driven manner, are the right macroscopic ``observables", i.e., 
the variables which parametrize the macroscale dynamical behavior of the system, thus enabling the construction of a reduced macroscale model.

Such manifold learning algorithms were initially applied to synthetic data sets, to illustrate
their geometric properties and flexibility \cite{coifman2005geometric, nadler2006diffusion}.
%
More recently they have been applied to experimental as well as simulation data,
enabled by advances in data representation (observers) and
metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally, rohrdanz2011determination, talmon2015manifold}.
%
A nontrivial shortcoming of these methods is that, from a geometric perspective,
not all eigenvectors are guaranteed to parametrize unique directions within the data;
some eigenvectors are ``repeated eigendirections'' which describe the same coordinate in the intrinsic geometry
of the data.
%
Identifying these eigenvectors is critical for obtaining a
parametrization of the system which captures the true dimensionality of the macroscale dynamics.
%
This task is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{gerber2007robust}.

In this paper, we propose an algorithm to automatically identify such unique eigendirections using local linear regression \cite{wasserman2006all}.
%
We first demonstrate and validate our algorithm on synthetic examples, where a closed-form solution for the
eigenvectors and eigenvalues of the Laplace operator is known.
%
We then consider a simulation data set from a stochastic dynamical system modeling
cellular chemotaxis \cite{othmer1988models}.
%
Recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections,
provide a data analysis pipeline which successfully analyzes this simulation in a purely data-driven manner.
%
We will show that this pipeline allows us to detect changes in the dimensionality of the underlying macroscale dynamics, which are
related to changes in the regime/mode of the system.

@@

Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ state observations sampled from an evolving autonomous dynamical system.
%
We assume the $n$-dimensional observations $z(i)$ lie on a $d$-dimensional manifold $\mathcal{M}_d$, where $d < n$.
%
We therefore consider the problem of parametrizing the continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$ from data.
%
For linear hyperplanes, the principal axes parametrize the manifold.
%
However, in the case when the manifold is nonlinear, a set of appropriate coordinates is not readily apparent.
%
By using a particular manifold learning technique based on the construction of a Laplace operator, namely, {\em diffusion maps}, we will show how to extract a $d$-dimensional parametrization of the observations which is consistent with the geometry of the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_cnts1}
\includegraphics[width=\textwidth]{strip_cnts2}
\caption{}
\label{subfig:strip_efuncs}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\includegraphics[width=\textwidth]{strip_discrete4}
\caption{}
\label{subfig:strip_evecs_uniform}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_nonuniform1}
\includegraphics[width=\textwidth]{strip_nonuniform2}
\caption{}
\label{subfig:strip_evecs_nonuniform}
\end{subfigure}
%
\caption{(a) Two-dimensional continuous strip colored by the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$. (b) Data, uniformly sampled from a two-dimensional strip, colored by the first and fourth (non-trivial) eigenvectors of the discrete Laplacian. (c) Data, sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the first and third (non-trivial) eigenvectors of the discrete Laplacian. Note that in all cases we uncover parametrizations which are one-to-one with $z_1$ and $z_2$ on the domain.}
\end{figure}

In recent years it has been noted that the eigenfunctions of the \emph{continuous}
Laplace-Beltrami operator provide ``good" coordinates for a manifold \cite{jones2008}.
%
We start by considering such a continuous setting, where rigorous analysis is possible for specific examples.
%
Using such an example, we will demonstrate that the existence of repeated eigendirections is
inherent to the manifold learning setup based on Laplace operators,
and that the identification of the unique eigendirections requires some additional effort.
%
The fact that these challenges arise in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for data analysis.

To review and illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$.
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}.
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and
$\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates,
respectively, on the domain, and therefore provide a parametrization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide an estimate of the
relative magnitude of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases,
the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases
(this will be discussed further in Section~\ref{sec:relative_lengths}).
%
The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper.
%
First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions,
and a proper parameterization of the manifold is not necessarily given by the $d$ eigenfunctions associated with the
smallest $d$ eigenvalues.
% TODO: address this in later section
%
Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any
additional directions intrinsic to the strip geometry; we will refer to these as ``repeated eigendirections,''
and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
%
Clearly, this problem of repeated eigendirections also arises for more curved of more than two variables.
%
We note that, although the eigenfunctions can only be written analytically for very special cases,
it has been observed empirically that they often provide appropriate coordinates
to usefully parametrize more complicated, nonlinear manifolds.
%

@@

Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we first construct the weight matrix $\mathbf{W} \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
\mathbf{W}_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations.
%
The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{rohrdanz2011determination, coifman2008graph};
we often take $\epsilon$ to be the median of the pairwise distances between the data points. 
%
The underlying assumption is that, even in cases where the samples embedded in the high dimensional space lie on a highly nonlinear manifold, within $\epsilon$-local neighborhoods, the Euclidean distance respects the tangent plane to the manifold and thus locally conveys meaningful relationships,. 
%
We then construct the diagonal matrix $\mathbf{D} \in \mathbb{R}^{m \times m}$, with $\mathbf{D}_{ii} = \sum_j \mathbf{W}_{ij}$,
and form the matrix $\widetilde{\mathbf{W}} = \mathbf{D}^{-\alpha} \mathbf{W} \mathbf{D}^{-\alpha}$, where $0 \le \alpha \le 1$.
%
Next, we construct the diagonal matrix $\widetilde{\mathbf{D}} \in \mathbb{R}^{m \times m}$, with $\widetilde{\mathbf{D}}_{ii} = \sum_j \widetilde{\mathbf{W}}_{ij}$, and the matrix $\mathbf{A}  = \widetilde{\mathbf{D}}^{-1} \widetilde{\mathbf{W}}.$

If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then, for $\epsilon \rightarrow 0$ and $m \rightarrow \infty$ (with the appropriate rates), the discrete matrix converges to the following continuous limit operators with Neumann boundary conditions (as discussed in the previous section) \cite{coifman2006geometric}
\begin{align} 
\label{eq:limiting_op1}
\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - 2\nabla U \cdot \nabla \phi, &&\alpha = 0 \\
\label{eq:limiting_op2}
\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, &&\alpha = 1/2 \\
\label{eq:limiting_op3}
\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi, &&\alpha = 1
\end{align}
where $U = - \log q$.
%
The different limit operators, depending on the choice of $\alpha$, imply that nonuniform
sampling on the manifold may have different effects via the potential $U$ in \eqref{eq:limiting_op1}--\eqref{eq:limiting_op3}.
%
In particular, by setting $\alpha=1$, one can factor out the density effects in the
weight matrix, and the discrete Laplacian matrix approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$.
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $\mathbf{A}$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $\mathbf{A}$ are related to the eigenvalues of the continuous operator by
\begin{equation} \label{eq:evals_relationship}
\mu_k = \exp \left( -\frac{\epsilon^2}{4} \tilde{\mu}_{k_1, k_2}  \right).
\end{equation}

@@

From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$
provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.
%
From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of
the corresponding eigenvalues implies that these two eigenvectors are guaranteed to
appear before any coupled or repeated eigendirections.
%
However, these eigenvectors are {\em not} guaranteed to appear as {\em the first two} (non-trivial) eigenvectors,
as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
%
We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold are unknown and cannot easily be obtained from the coordinates of the original data, so that identifying which eigenvectors correspond to unique eigendirections is nontrivial.

@@

Common practice is to order the eigenvectors by the magnitude of the corresponding eigenvalues,
assuming that the leading eigenvectors provide a parametrization of the underlying manifold.
%
However, as discussed in the previous section, some eigenvectors are higher harmonics of
previous eigenvectors and do not describe new directions in the data set \cite{gerber2007robust}.
%
For the case where $\mathcal{M}_d$ is a $2$-dimensional strip with edge lengths $L_1  > L_2$, recall
that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left(  {\pi z_1}/{L_1} \right)$ and
$\tilde{\phi}_{0,1} = \cos \left(  {\pi z_2}/{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to
correspond to the two smallest (non-trivial) eigenvalues (the smallest eigenvalue will always be $\tilde{\mu}_{0,0} = 0$ and
correspond to a constant eigenfunction; these eigenvalues are related to the eigenvalues $\mu_k$ of the discrete
operator via \eqref{eq:evals_relationship}).
%
In fact, if $L_1 > 2 L_2$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the
second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues)
will be a ``repeated eigendirection" of the first and still parametrize $z_1$ (see Figure~\ref{fig:strip_harmonics}).
%
Automatic detection of which eigenvectors are harmonics of each other is clearly useful.
%
Utilizing only the eigenvectors $\phi_i$ which correspond to unique eigendirections yields
the most parsimonious representation of the data.
%
We will show how we can automatically detect these eigenvectors to obtain a meaningful representation of the data,
and that using only these eigenvectors yields a reduced embedding which is equivalent to the standard diffusion maps embedding.
%
Furthermore, when the data is uniformly sampled from the underlying manifold, the corresponding eigenvalues provide us with an estimate of the relative lengths of the data set along these unique eigendirections.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\end{subfigure}
%
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete2}
\end{subfigure}
%
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete3}
\end{subfigure}
%
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete4}
\end{subfigure}
\caption{ Data, uniformly sampled from a two-dimensional strip, colored by the first four (non-trivial) eigenvectors from diffusion maps. Note that the first and fourth eigenvectors are one-to-one with $z_1$ and $z_2$, respectively. However, the second and third eigenvectors are higher harmonics of the first eigenvector and do not capture any additional structure within the data set. }
\label{fig:strip_harmonics}
\end{figure}
